Supplementary Materials

### 2023-07-07 manuscript/manuscript.tex
paper changes

\RequirePackage[bibstyle=authoryear,citestyle=authoryear-comp,
                date=year,
                maxbibnames=9,maxnames=5,maxcitenames=2,
                backend=biber,uniquelist=false,uniquename=false,
                % style=apa,
                sorting=nyt,
                % sorting=,
                hyperref=true]{biblatex}
\usepackage{color}
\usepackage{nicefrac}

@@

%\title{Towards a Complete Model of the Negative Selection Process in Humans}
%\title{An Evolutionary Quantitative Genetic Model of the Negative Selection Process in Humans}
\title{A Quantitative Genetic Linked Selection Model fit to the Negative Selection Process in Humans}

@@

\begin{abstract} 
% TODO: emphasize connection to nearly neutral theory
\end{abstract}

@@

However, across all of our annotation models, new deleterious mutations in
conserved feature classes were consistently estimated to have strongly
deleterious effects (Figure \ref{fig:figure-3}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. This includes features such as
coding sequences, PhastCons regions, or the top 6\% of basepairs with the most
pathogenic CADD scores. However, we find that there is considerable uncertainty
in how strong selection is against the most deleterious mutations. For example,
in the CADD 6\% full-track model, new mutations are estimated to have a
selection coefficients of $s = 10^{-1}$ with 42\% chance, $s=10^{-2}$ with
7.8\% chance, and $s=10^{-3}$ with 41\% chance. This creates a bimodal DFE
(Figure \ref{fig:figure-3}A, second row, left column), which is mirrored across
the other CADD models. Moreover, these DFE estimates suggest a very high
average selection coefficient across CADD 6\% regions of $\widehat{\bar{s}}
\approx 0.043$ for YRI.

Given that previous methods have typically not estimated the DFE up to
selection coefficients of $s=10^{-1}$, we ran additional ``constrained grid"
models using a DFE only up to $s=10^{-2}$. Across all populations, these
constrained grid models fit the data less well than the default grid by about
one percentage point less in $R_\text{LOCO}^2$ (see Supplementary Table
\ref{supp:tbl-r2}). As expected, the constrained grid models distributed mass
onto the next strongest selection coefficient, $s=10^{-2}$, and the average
selection was reduced about six-fold to $\widehat{\bar{s}} \approx 0.0079$ for
YRI. While the DFE may truly be bimodal for highly-conserved regions and have a
very high average selection coefficient, a likely alternative is that these
results reflect a non-identifiability issue between selective effects of
$s=10^{-1}$ and $s=10^{-2}$. We will refer to this as the
\emph{strong-selection non-identifiability} issue.

@@

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_5.pdf} 
    \caption{(A) CADD 6\% full-track }
  \label{fig:figure-5}
\end{figure}

While we have previously shown that out-sample predicted diversity matches
observed diversity (Section XXX), we also investigated whether there were
regions of the genome that were particularly poor-fitting. Comparing predicted
against observed diversity along chromosomes (see Supplementary Materials
Section \ref{supp:chrom-fit}) we overall find close correspondence consistent
with the high $R_\text{LOCO}^2$. However, we find a few large (tens of
megabases) regions with systematically poorer fit (Supplementary Materials
Figure \ref{supp:spatial-residuals}). In Figure \ref{fig:figure-5}A we see one
such region from XXX to XXX. Interestingly, predicted diversity closely follows
the peaks and troughs of this region. We note that a smaller segment within
this region has been found by a genome-wide scan for associative overdominance
\parencite{Gilbert2020-aw}.

@@

\subsection*{Predicted Substitution Rates, and Fitness Variance}

@@

    \caption{Evidence that negative selection signal remains. (A) Residuals
        from the YRI CADD 6\% model plotted against the average LoF selection
        coefficient across genes in a megabase window (estimated from
        \cite{Agarwal2023-un}). Our model is systematically over predicting
        diversity in regions estimated to be sensitive to LoF variants. The
        ordinary least squares (OLS) estimate suggests that 2.1\% of the
        variance in residuals can be explained by reductions caused by the
        local sensitivity to LoF mutations. (B) Residuals from the YRI CADD 6\%
        model plotted against the fraction of CADD 6\% basepairs in the
        megabase window. Here, only 0.8\% of the variance is explained by
        remaining spatial variance in CADD 6\% regions (blue). Overlaid is the CADD
        2\% OLS curve (orange). }

@@

Since our B' method is based upon quantitative genetic models of linked
selection that explicitly incorporates the substitution process, fitting our
model to genomic data leads to predictions that can be explicitly tested. In
particular, there are two key predicted quantities: (1) the substitution rates
per segment, and (3) the fitness variance per segment.

Each of our maximum likelihood estimates $\widehat{\Psi}$ imply other
corresponding

Our method accommodates weak selection by modeling the fitness variance in each
segment under an equilibrium negative selection process. This fitness variance
is found by simultaneously solving for the per-generation substitution rate $R$
and the draft-effective population size $N_d$ for each segment. When our model
is fit to genome-wide diversity data, the parameter estimates imply predictions
of the deleterious substitution rate per basepair, per generation, for each
feature. Our method only considers within-species pairwise diversity, so these
predicted substitution rates constitute a test that our model is making
reasonable out-sample predictions of the substitution rates across features.

We focus specifically on the predictions from our feature-based annotation
models, since we can readily estimate human substitution distances for each
feature to compare to our predictions.

The estimated substitution distance should in theory equal the product of the
substitution rate per site, per year and the branch length in years.

When comparing our predicted mutation rates to observed substitution distances,
we must account for three major sources of uncertainty.

First, the split time between humans and chimpanzees is unknown, with estimates
ranging from 6Mya to 12Mya (XXX). Second, our predictions are in sites per
generation, and the appropriate human generation time is unknown. Third, the
appropriate mutation rate to use is debated, since there may have been a
mutational slow down on the human lineage (XXX), which could explain why
phylogenetic- and pedigree-based mutation rate estimates differ by roughly a
factor of two (XXX). Finally, we are only considering modeling the deleterious
substitution rate. In reality, some fraction of observed substitutions fixed
because they were beneficial.

For a particular feature $f$, the estimated per-site substitution distance
$S_f$ along a branch (which accounts for back mutations) should correspond to
$S_f = T_g r_f$, where $T_g$ is the time in generations and $r_f$ is the
substitution rate per basepair, per generation of that feature.

Each feature is
composed of basepairs with varying selection coefficients distributed according
to the feature-specific DFE vector $\mathbf{w}_f$ and corresponding vector of
substitution rates per selection coefficient, $\mathbf{r}_f$.

Our estimate of a feature's DFE
is $\widehat{\mathbf{w}}_f$.

We calculate the

\begin{align}
    \label{eq:sub}
    r_f = \mu \int_0^1 p_F(s) ds
\end{align}

Solving the XXX equations, we end up with predictions $\widehat{r}_f$ for the
per-basepair, per-generation substitution rate for each feature. These are
calibrated to the MLE mutation rate estimate $\widehat{\mu}_\text{del}$, which
as mentioned, excludes the contribution of lethal mutations which act purely to
reduce $N_e$ and thus only impact $\pi_0$. Using Equation \eqref{eq:sub} we can
re-calibrate these estimates against other mutation rates.

Because of it's greater interpretabilty, we initially focus on the substitution
rate estimates from the Feature Priority model even though it had a poorer-fit
and high mutation rate estimate.

We can also re-calibrate this model to an alternate mutation rate.

% - Table s5 https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/genetics/206/1/10.1534_genetics.116.197145/8/files1.pdf?Expires=1686357613&Signature=cp9qoI2ChCb6GFZ74HI4eWcA~MQ4sHlWyeI3Zp49-5xEORzMfYlSRxzVvZlo8ZoGWDi0G3PnL00ZplM-~or4YIVm2uhkmqjf61AmC7RdZ1QTak4j0kFMQ0wAbM9pusiV3gIepTQHq2HKRKwgS~ypmv~0oqO5dDxaq9nhOWIBBmLAA49oHWuxUtnRe3sEYATwCwjIFEomhgJpg7vKGCF~87GdCV4JETW4PN2VvD4hsu2AVRd8yA~vUJvsXfhWwcElDeY4fsgpprl1lC3XGjJCnqdwbzRGPV5LNx5XEKLptcZtLeLcOmFJCvmVLe0bewJ7Vta0TSQM7T9AuklompjPPw__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA

Like other methods, our method divides the putatively conserved basepairs up

Generally, when the putatively conserved region is larger, there is a tradeoff
in two dimensions. First, more bases are conserved according to the same DFE,
so the total mutational load increases. This causes the tradeoff between
mutation rate estimate and CADD fraction we observe across both our B' model
and the classic B when using sparse tracks (Figure 2XXX).

We find that mutation rate estimates are fairly sensitive to model choice
(Figure 2XXX).

Since B' has a U-shaped relationship with the selection coefficient, we
hypothesized that our model may reveal true identifiability issues in
estimating the DFE that do not occur under the classic B model. We find that
this is not the case for our models fit at the megabase-scale. The DFE
estimates for our CADD6 and CADD8 models are consistent (Figure XXX), and both
produce similar mutation rates.

While we do not comprehensively fit all of our model to other spatial scales,
we

However, while a given reduction level may be explained by either a weak or
strong selection coefficient, the spatial extent of these reductions caused by
weak selection is much narrower than under strong selection.

We see evidence this is the case in several of our models, where
genome-wide diversity fits equally well between models with mutation rates

Indeed, we see this by comparing parameter estimates of the CADD6
sparse-track model using classic BGS B and our B' method.

(which mimics the best-fitting model of
\cite{Murphy2022-sj})

(see also
Appendix 1, Figure 26A of \cite{Murphy2022-sj})

correlation between s and feature

Overall, we find evidence that negative selection models are very sensitive to
the input track. Because inference based on our B' method accommodates weaker
selection, the primary models we fit here more robust to misspecification of
the regions

@@

>>>>

\RequirePackage[bibstyle=authoryear,citestyle=authoryear-comp,
                date=year,
                maxbibnames=9,maxnames=5,maxcitenames=2,
                backend=biber,uniquelist=false,uniquename=false,
                % style=apa,
                sorting=nyt,
                % sorting=,
                hyperref=true]{biblatex}
\RequirePackage{color}
\RequirePackage{nicefrac}

@@

%\title{Towards a Complete Model of the Negative Selection Process in Humans}
%\title{An Evolutionary Quantitative Genetic Model of the Negative Selection Process in Humans}
\title{A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans}

@@

\begin{abstract} 
% TODO: emphasize connection to nearly neutral theory

    % uniting quant gen/fitness variation, linked selection, and substitution
    % The interesting thing about connecting all these models is that we might
    % have hope of calibrating these models to the true process.
\end{abstract}

@@

However, across all of our annotation models, new deleterious mutations in
conserved feature classes were consistently estimated to have strongly
deleterious effects (Figure \ref{fig:figure-3}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. This includes features such as
coding sequences, PhastCons regions, or the top 6\% of basepairs with the most
pathogenic CADD scores. However, we find that there is considerable uncertainty
in how strong selection is against the most deleterious mutations. For example,
in the CADD 6\% full-track model, new mutations are estimated to have a
selection coefficients of $s = 10^{-1}$ with 42\% chance, $s=10^{-2}$ with
7.8\% chance, and $s=10^{-3}$ with 41\% chance. This creates a bimodal DFE
(Figure \ref{fig:figure-3}A, second row, left column), which is mirrored across
the other CADD models. Moreover, these DFE estimates suggest a very high
average selection coefficient across CADD 6\% regions of ${\bar{s}}
\approx 0.043$ for YRI.

Given that previous methods have typically not estimated the DFE up to
selection coefficients of $s=10^{-1}$, we ran additional ``constrained grid"
models using a DFE only up to $s=10^{-2}$. Across all populations, these
constrained grid models fit the data slightly less well than the default grid
by about one percentage point less in $R_\text{LOCO}^2$ (see Supplementary
Table \ref{supp:tbl-r2}). As expected, the constrained grid models distributed
mass onto the next strongest selection coefficient, $s=10^{-2}$, and the
average selection was reduced about six-fold to ${\bar{s}} \approx 0.0079$ for
YRI. Our constrained-grid DFE estimates for the Feature Priority model are
qualitatively consistent with the U-shaped DFEs found for amino acids through
Poisson Random Field method \parencite{Boyko2008-tj}. While the DFE may truly
be bimodal for highly-conserved regions and have a very high average selection
coefficient, a likely alternative is that these results reflect a
non-identifiability issue between selective effects of $s=10^{-1}$ and
$s=10^{-2}$. We will refer to this as the \emph{strong-selection
non-identifiability} issue.

@@

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_5.pdf}

    \caption{(A) Observed and predicted diversity under the CADD 6\% full-track
        model. Once scaled by average diversity, populations differ little in
        their predicted diversity levels (colors). Additionally, we show
        summaries of CADD density and recombination rate along the chromosome
        below. (B) Predicted $B$ and observed $\pi$ for each window. The
        residual variance is in the vicinity of the levels expected from the
        $R_\text{LOCO}^2$ calculation. (C) Evidence that negative selection
    signal remains in CADD 6\% residuals (YRI shown). Residuals are plotted
against the average LoF selection coefficient across genes in a megabase window
(estimated from \cite{Agarwal2023-un}).}

  \label{fig:figure-5}
\end{figure}

While we have previously shown that out-sample predicted diversity matches
observed diversity (Section XXX), we also investigated whether there were
regions of the genome that were particularly poor-fitting. Comparing predicted
against observed diversity along chromosomes (see Supplementary Materials
Section \ref{supp:chrom-fit}) we find a close correspondence consistent with
the high $R_\text{LOCO}^2$. However, we find a few large (tens of megabases)
regions with systematically poorer fit (Supplementary Materials Figure
\ref{supp:spatial-residuals}). In Figure \ref{fig:figure-5}A we see one such
region on the short arm of chromosome 2, from 30 Mbp to 60Mbp. Interestingly,
predicted diversity closely follows the peaks and troughs of this region. We
note that a small region within this stretch had been found by a genome-wide
scan for associative overdominance \parencite{Gilbert2020-aw}.

@@

Earlier, we found that the goodness-of-fit and DFE estimates of our
feature-based annotation models are sensitive to how basepairs are grouped into
feature classes. Our models are in essence fitting a single distribution across
sites within a feature class, yet there may be local heterogeneity among sites
that is poorly captured by this approach. We investigated this possibility by
looking for lingering selection signal in our model residuals. First, we
inspected whether there was a relationship between the fraction of CADD 2\% and
6\% basepairs and the residual across megabase windows (Supplementary Figure
\label{suppfig:resid-cadd}), finding a negative significant relationship in
both cases. Moreover, our model over-predicted diversity in windows containing
more CADD 2\% basepairs moreso than CADD 6\%, consistent with heterogeneity in
site pathogenicity being poorly fit by our model. However, the total residual
variance explained is $R^2 = 0.4\%$ and $R^2 = 0.8\%$ for the CADD 2\% and 6\%
tracks respectively, suggesting only modest amount of selection signal remains.

We further tested for remaining selection signal in our CADD 6\% model
residuals by using gene-specific estimates of the fitness cost of
loss-of-function (LoF) mutations from \textcite{Agarwal2023-un}. These
estimates are based on an Approximate Bayesian Computation approach that
estimates the posterior distribution over LoF fitness costs from the observed
dearth of LoF mutations per gene, and thus is an independent methodological
approach than ours. We averaged the estimated LoF fitness costs across genes
for each of our megabase windows, and plotted our residuals against these
average LoF fitness costs. Contrary to the weak CADD residual signal described
above, we find evidence of a fairly strong relationship between our residuals
and average LoF fitness cost (Figure \ref{fig:figure-5}C; $R^2 = 2.1\%$,
p-value $1.27 \times 10^{-10}$). In other words, roughly 2\% of the variance in
these residuals is explained by the average fitness costs of LoF mutations in
the window. Consequently, our model over-predicts diversity by about
$\nicefrac{\sigma}{2}$ or more in windows harboring the top 1.7\% most
LoF-intolerant genes.

\subsection*{Predicted Substitution Rate and Additive Fitness Variance}

@@

    \caption{}

@@

Since our B' method is based upon quantitative genetic models of linked
selection that explicitly incorporates the substitution process, fitting our
model to genomic data leads to predictions: (1) the substitution rates per
segment, (2) the fitness variance per segment, and (3) the total number of
strongly deleterious mutations per diploid genome per generation. Given
predicted substitution rates for each feature can be compared to estimated
sequence divergence between species, this offers a qualitative check our model
is producing sensible results.

To compare our models' predicted substitution rates against observed
divergence, we estimated sequence divergence on the human lineage using a
multiple alignment of five primates for each feature in our feature-based
models (Methods XXX). Since we have observed that mutation rate estimates from
our method can be unreliable (and simulation results confirm this), we
predicted substitution rates under a fixed mutation rate of $\mu = 1.5 \times
10^{-8}$, rather than the maximum likelihood estimates. Unfortunately, a
careful comparison between our predictions and observed divergence rates is
hindered by considerable uncertainty in the rate of the molecular clock,
generation times, and the human-chimpanzee divergence time. We assume a
generation time of 29 years, and calculate the sequence divergence implied by
our predicted substitution rates over a range of divergence times, from 6 Mya
to 12 Mya \parencite{Moorjani2016-tb,Nachman2000-te,Yi2002-pw,Steiper2006-xx}.

Our predictions are qualitatively consistent with the observed divergence along
the human lineage (Figure \ref{fig:figure-6}). As expected, the predicted
substitution rate in features unlikely to be under selective constraint is very
close to the mutation rate. Throughout, we report our substitution rates as a
percent relative to the total mutation rate, $\mu$ (here fixed to $1.5 \times
10^{-8}$). In our PhastCons Priority model, introns and UTRs are estimated to
have have substitution rates that are $93.6\%$ of the mutation rate, and the
``other" feature is estimated at $97.8\%$. When we use the constrained grid,
these change to $97.85\%$ and $99.98\%$ respectively. By contrast, our
predicted substitution rates for PhastCons regions are much lower: $0.48\%$ for
the default grid, and $1.65\%$ for the constrained grid. This is likely due to
the constrained grid placing more of the DFE mass on weaker selection
coefficients, which in turn have a higher chance of drifting up to fixation.

Finally, we entertain two more speculative predictions. First, the equilibrium
additive genic fitness variance. Our PhastCons Priority model DFE estimates and
negative selection model imply a $\widehat{V}_a = 0.45$ for the default grid,
and $\widehat{V}_a = 0.040$ for the constrained grid. This is a considerable
level of uncertainty---over an order of magnitude---due to a difference in the
selection grid. There are no other known direct estimates in humans of this
quantity for comparison. When $2N \gg 1$, theory suggests $V_a \approx
U_D\bar{s}$ where $U_D$ is the diploid rate of strongly deleterious mutations.
Thus, the uncertainty is a byproduct of the different average estimates of
$\bar{s}$ across the alternate grids. Since both this work and
\textcite{Murphy2022-sj} find the single-feature CADD 6\% track fits well, we
can use this to gain intuition through some rough calculations. In this case,
$U_D \approx 6\% \times 2 \times 3 \times 10^{9} \mu \approx 3.6 \times 10^{8}
\mu$. For $\mu \approx 1.5 \times 10^{-8}$, $U_D \approx 5.2$. Under the
constrained grid, ${\bar{s}} \approx 0.0079$ and thus $V_a \approx 0.043$; with
the default grid $\bar{s} \approx 0.043$, and $V_a \approx 0.23$. Often fitness
reductions of $\approx 1\%$ are assumed for the most highly-conserved regions
(XXX, Keightly, Lynch); for comparison, this leads to $V_a \approx 0.054$.

XXX HERE  USE CADD PREDICTIONS

By theory, this relative fitness (scaled by the mean squared) is
bounded from above by the Index of Selection, defined as the theoretic relative
fitness variance if fitness were perfectly heritable \parencite{Crow1958-pc}.
Based on rough calculation by Crow, this could range between 0.982-3.286 in
humans. However, this is based on modern-day mortality figures that may not
extrapolate to the timescales that determine genome-wide patterns of diversity.

Second, we can calculate the reduction in relative fitness per generation due
to deleterious substitutions. Since \textcite{Muller1950-ta}, there has been
considerable interest in whether mildly deleterious mutations, perhaps due to
the relaxation of selection pressures, could accumulate and fix at a faster
pace in humans.

Here, the selection grid choice has little impact
on the estimates: we calculate that across the genome, the loss in relative
fitness per generation is 0.00160\% for the default grid, and 0.00119\% for the
constrained grid.

@@

- parametric DFEs are trash

### 2023-07-07 manuscript/supplementary.tex
paper changes

The full list of samples is available in TSV format in the GitHub repository
(\path{data/h1kg/yri_samples.tsv}).

XXX

@@

gVCFs were filtered using a custom Python tool, \texttt{gvcf2counts.py} (in
\path{tools/gvcf2counts.py}), which reads the gVCFs, filters them according to
the criteria below, and outputs a Numpy \path{.npy} file of reference and
alternative allele counts for each chromosome (hereafter, ``allele counts
matrices").

@@

Generally, fitting procedures act to minimize the squared deviation between
predictors and observed values, which drives $\cov_i(\varepsilon_i,
\widehat{B}_i)$ to zero XXX. \textcite{Murphy2022-sj} suggest  XXX

@@

\begin{figure}[htbp] \centering
    \includegraphics[width=0.5\textwidth]{figures/supplementary/subfigure_R2_scale.pdf}
    \caption{XXX}
  \label{suppfig:coal_r2_scale}
\end{figure}

@@

>>>>

@@

The gVCFs were filtered using a custom Python tool, \texttt{gvcf2counts.py} (in
\path{tools/gvcf2counts.py}), which reads the gVCFs, filters them according to
the criteria below, and outputs a Numpy \path{.npy} file of reference and
alternative allele counts for each chromosome (hereafter, ``allele counts
matrices").

@@

@@

% \begin{figure}[htbp] \centering
%     \includegraphics[width=0.5\textwidth]{figures/supplementary/subfigure_R2_scale.pdf}

%     \caption{How the window size effects our theoretic $R_\text{coal}^2$ versus
%         our CADD 6\% models classic BGS (B) and new B' methods at two scales (1
%         Mbp and 100 kbp). Additionally, we have overlaid the CADD 6\%
%         out-sample $R_{OS}^2$ estimates across many window sizes from
%     \textcite{Murphy2022-sj}. We note that the theoretic $R_\text{coal}^2$
% quickly declines to zero since the coalescence and mutational noise quickly
% begins the linked selection signal.}

%   \label{suppfig:coal_r2_scale}
% \end{figure}

@@

\begin{figure}[htbp]
  \label{suppfig:resid-cadd}
  \centering
  \includegraphics[width=0.6\textwidth]{figures/supplementary/cadd_residual.pdf}
  \caption{Residuals from the CADD 6\% sparse track plotted against the fraction of
  basepairs in a window annotated by a CADD region.}
\end{figure}
